\section{What is Dimple?}

Dimple is an open-source API for probabilistic modeling and inference. Dimple allows the user to specify probabilistic models in the form of graphical models (factor graphs, Bayesian networks, or Markov networks), and performs inference on the model using a variety of supported algorithms.

Probabilistic graphical models unify a great number of models from machine learning, statistical text processing, vision, bioinformatics, and many other fields concerned with the analysis and understanding of noisy, incomplete, or inconsistent data. Graphical models reduce the complexity inherent in complex statistical models by dividing them into a series of logically (and statistically) independent components. By factoring the problem into sub-problems with known and simple interdependencies, and by adopting a common language to describe each sub-problem, one can considerably simplify the task of creating complex probabilistic models.  A brief tutorial on graphical models can be found in Appendix~\ref{sec:IntroductionToFactorGraphs}.

An important attribute of Dimple is that it allows the user to construct probabilistic models in a form that is largely independent of the algorithm used to perform inference on the model. This modular architecture benefits those who create probabilistic models by freeing them from the complexities of the inference algorithms, and it benefits those who develop new inference algorithms by allowing these algorithms to be implemented independently from any particular model or application.

Key features of Dimple:
\begin{itemize}
\item Supports both undirected and directed graphs.
\item Supports a variety of \emph{solvers} for performing inference, including sum-product and Gaussian belief propagation (BP), min-sum BP, particle BP, discrete junction tree, linear programming (LP), and Gibbs sampling.
\item Supports both discrete and continuous variables.
\item Supports arbitrary factor functions as well as a growing library of standard distributions and mathematical functions.
\item Supports nested graphs.
\item Supports rolled-up graphs (repeated HMM-like structures).
\item Growing support for parameter estimation (including the EM algorithm).
\item Supports both MATLAB\footnote{MATLAB is a registered trademark of The Mathworks, Inc.} and Java\footnote{Java is a registered trademark of Oracle and/or its affiliates.} API.
\end{itemize}

\ifmatlab
Using MATLAB to program Dimple requires an existing licensed version of MATLAB, of version at least 2013b.
\fi
